---
title: "Installation of R & RStudio"
output: 
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


<br/>



`r fontawesome::fa_i("hand-point-right")`&nbsp;
**Please install the latest versions of `r fontawesome::fa_i("r-project")`
and RStudio.**



<br/>



## Installation



### `r fontawesome::fa_i("windows")` &nbsp;Windows



1. Download `r fontawesome::fa_i("r-project")` at:<br/>
[https://cran.r-project.org/](https://cran.r-project.org/)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _Download R for Windows_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _base_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Click on _Download R 4.2.0 for Windows_
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _R-4.2.0-win.exe_ and follow instructions



<br/>



2. Download **`RStudio Desktop`** at:<br/>
[https://www.rstudio.com/products/rstudio/download/#download](https://www.rstudio.com/products/rstudio/download/#download)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Download the version for Windows 10/11 (the **.exe** and not the **.zip**)
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _RStudio-2022.02.2-485.exe_ and follow instructions
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Check if **`RStudio`** has found `r fontawesome::fa_i("r-project")`. 
Launch **`RStudio`** and run:

```{r echo=TRUE, eval=FALSE}
R.version.string
## [1] "R version 4.2.0 (2022-04-22) -- "Vigorous Calisthenics"
```


<br/>



3. Download **`Rtools4.2`** at:<br/>
[https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html](https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html)

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Download the 64-bit version for Windows
<br/>&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;`r fontawesome::fa_i("caret-right")` &nbsp;
Open _rtools42-5253-5107.exe_ and follow instructions

The installation of **`Rtools4.2`** is strongly recommended to install some packages 
from sources. 
**`Rtools4.2`** contains in particular the utility `make`.

Restart **`RStudio`** and check that `r fontawesome::fa_i("r-project")` finds the utility `make`:

```{r echo=TRUE, eval=FALSE}
Sys.which("make")
## "C:\\rtools42\\usr\\bin\\make.exe"
```

Now you can install packages from sources:


```{r echo=TRUE, eval=FALSE}
install.packages("jsonlite", type = "source")
```


<br/>



### `r fontawesome::fa_i("apple")` &nbsp;macOS


Th easiest way to set up your environment is by using the package manager for macOS
[**`Homebrew`**](https://brew.sh). With this tool you can install (and update) your
software directly from the terminal. First you need to install the Apple software
_Command Line Tools for Xcode_.

Open a [terminal](https://support.apple.com/en-gb/guide/terminal/apd5265185d-f365-44cb-8b09-71a064a42125/mac), 
run this line and accept the license:

```sh
## Install Xcode Command Line Tools ----
sudo xcode-select --install
```

Then you can install **`Homebrew`**:

```sh
## Install Homebrew ----
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

## Check installation ----
brew --version
## Homebrew 3.4.10

## Update repositories (list of available software) ----
brew update
```


Follow these lines to install your environment:

```sh
## Install utilities for R ----
brew install --cask xquartz     ## Graphical window system
brew install gfortran           ## FORTRAN compiler for GCC

## Install R ----
brew install --cask r

## Install RStudio Desktop ----
brew install --cask rstudio
```

Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
R.version.string
## "R version 4.2.0 (2022-04-22) -- "Vigorous Calisthenics"

Sys.which("make")
##            make 
## "/usr/bin/make"
```

<br/>

To keep your system up-to-date, run regularly:


```sh
## Update environment ----
brew update && brew upgrade --greedy
```


<br/>



### `r fontawesome::fa_i("ubuntu")`&nbsp;Ubuntu


First check your version of Ubuntu in a terminal:


```sh
## Check Ubuntu version ----
cat /etc/issue
## Ubuntu 22.04 LTS
```

<br/>

| Release   | Release name    | Repository URL                                               |
|:---------:|:----------------|:-------------------------------------------------------------|
| 22.04 LTS | Jammy Jellyfish | https://cloud.r-project.org/bin/linux/ubuntu jammy-cran40/   |
| 21.10     | Impish Indri    | https://cloud.r-project.org/bin/linux/ubuntu impish-cran40/  |
| 21.04     | Hirsute Hippo   | https://cloud.r-project.org/bin/linux/ubuntu hirsute-cran40/ |
| 20.04 LTS | Focal Fossa     | https://cloud.r-project.org/bin/linux/ubuntu focal-cran40/   |
| 18.04 LTS | Bionic Beaver   | https://cloud.r-project.org/bin/linux/ubuntu bionic-cran40/  |
| 16.04 LTS | Xenial Xerus    | https://cloud.r-project.org/bin/linux/ubuntu xenial-cran40/  |
: Table 1. List of CRAN repositories for different versions of Ubuntu

<br/>

Install `r fontawesome::fa_i("r-project")` by following these steps 
(change the URL below by the one specific to your version of Ubuntu, cf. Table 1) :

```sh
## Install APT utilities ----
sudo apt install wget software-properties-common

## Add CRAN GPG key ----
wget -qO- https://cloud.r-project.org/bin/linux/ubuntu/marutter_pubkey.asc | sudo tee -a /etc/apt/trusted.gpg.d/cran_ubuntu_key.asc

## Add CRAN repository to APT repositories list ----
sudo add-apt-repository 'deb https://cloud.r-project.org/bin/linux/ubuntu jammy-cran40/'

## Update packages list ----
sudo apt update

## Install R ----
sudo apt install r-base

## Install R (for development, recommended) ----
sudo apt install build-essential
sudo apt install r-base-dev
```


<br/>


To install **`RStudio Desktop`** visit this page: 
https://www.rstudio.com/products/rstudio/download/#download, download the
`.deb` file for Ubuntu 18+/Debian 10+ and install it. If you have an older version 
of Ubuntu (16.04), visit this page: https://www.rstudio.com/products/rstudio/older-versions/

Alternatively, you can install **`RStudio Desktop`** from the terminal: 

```sh
## Install wget utility ----
sudo apt install wget

## Download RStudio Desktop installer (Ubuntu >= 18.04) ----
wget https://download1.rstudio.org/desktop/bionic/amd64/rstudio-2022.02.2-485-amd64.deb

## Download RStudio Desktop installer (Ubuntu 16.04 only) ----
# wget https://download1.rstudio.org/desktop/xenial/amd64/rstudio-1.4.1106-amd64.deb

## Install RStudio Desktop ----
sudo dpkg -i rstudio-2022.02.2-485-amd64.deb
# sudo dpkg -i rstudio-1.4.1106-amd64.deb

## Remove installer ----
rm rstudio-2022.02.2-485-amd64.deb
# rm rstudio-1.4.1106-amd64.deb
```


<br/>


Open **`RStudio`** and check your installation:

```{r echo=TRUE, eval=FALSE}
R.version.string
## "R version 4.2.0 (2022-04-22) -- "Vigorous Calisthenics"

Sys.which("make")
##            make 
## "/usr/bin/make"
```

<br/>

To keep your system up-to-date, do not forget to frequently run:


```sh
## Update system ----
sudo apt update && sudo apt upgrade -y
```



<br/>




## For the workshop

You will need to install the [`rstan`](https://mc-stan.org/users/interfaces/rstan) package:


```{r echo=TRUE, eval=FALSE}
## CRAN package ----
install.packages("rstan")

## Check the installation ----
example(stan_model, package = "rstan", run.dontrun = TRUE)
```

If the example model fails, please read this [page](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#verifying-installation).




## Corrections {.appendix}


If you see mistakes or want to suggest changes, please 
[Create an issue](https://github.com/theodatasci/theodatasci.github.io/issues)
on the source repository.



## Reuse {.appendix}


The material of this website is licensed under Creative Commons Attribution 
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).
Source code is available at https://github.com/theodatasci/theodatasci.github.io/.



## Citation {.appendix}



Calcagno V, Fronhofer EA, Gounand I & Massol F (2022) 
Workshop FRB-CESAB & GdR TheoMoDive: Theory-driven Analysis for Ecological Data (Edition 2022). 
