---
title: "How to analyse a model"
author: "Isabelle Gounand"
date: "2024-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyzing the Rosenzweig  and MacArthur model

$\frac{dP}{dt}=r_0P(1-\frac{P}{K}) - \frac{aPH}{1+ahP}$

$\frac{dH}{dt}= - mH+ \epsilon\frac{aPH}{1+ahP}$

```{r rma_model,echo = FALSE}

library(deSolve)

# Rosenzweig and MacArthur model, function of the derivatives
rma_model <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dP = r0*P*(1-P/K) - a*P*H/(1+P*a*h)
    dH =  eps*a*P*H/(1+P*a*h) - m*H
    list(c(dP,dH))
  })
}


```

## Find the equilibria

Solve the system when:

$\frac{dP}{dt}=0$ 

AND 

$\frac{dH}{dt}=0$

That is express $P^*$ and $H^*$ with only the parameters when the derivatives of the system are null.
By hand when simple enough or with software doing symbolic calculus like Maxima (open source), Mathematica or Matlab (commercial).

We can study this by tracing the isoclines in the phase plan of P against H population densities, that is the criteria for which each of the equations equals zero. The equilibria are the intersection points within the feasible space:  $(P, H)>= 0$ 

```{r trace_isoclines}
# non-zero isocline functions 
zeroP = function(r0,K,a,h, P){r0/a*(1+P*a*h)*(1-P/K)}
zeroH = function(eps,a,h,m){m/(a*(eps-m*h))}


# Set the parameter values
r0 = 1
K = 3.5
a= 1.3
eps= 0.75
m=0.5
h=0.8
pars = c(r0 = r0, a = a, eps = eps, m=m, h=h,K=K)


# Set an array of P values
Ps = seq(0,K,by=0.1)


# calculate the isoclines for the given parameter set
H_zerogrowth = zeroH(eps,a,h,m)
P_zerogrowth = zeroP(r0,K,a,h,Ps)

colo = c("chartreuse3","darkorange3")


# plot isoclines
plot(NA,xlim = c(0,K*1.2),ylim = c(0,max(P_zerogrowth)*1.2),xlab="P",ylab="H",las=1,main = "Isoclines in phase plane")
abline(h=0,col=colo[2],lwd=2)
abline(v=c(0,H_zerogrowth),col=colo,lwd=2)
lines(Ps[P_zerogrowth>=0],P_zerogrowth[P_zerogrowth>=0],col=colo[1],lwd=2)

# Matrix of all equilibria
pstar = zeroH(eps,a,h,m)
hstar = zeroP(r0,K,a,h, pstar)
alleqs = matrix(c(0,0,K,0,pstar,hstar),ncol=2,byrow=T,dimnames=list(NULL,c("P","H")))

# Plot equilibria at intersection of H and P isoclines
points(alleqs[,"P"],alleqs[,"H"],lwd=2)
```


Here we get three possible equilibria:

$P^*=0$, $H^*=0$

$P^*=K$, $H^*=0$

$P^*=\frac{m}{a(\epsilon-hm)}$, $H^*=\frac{\epsilon(aKr_0(\epsilon-hm)-m)}{a^2K(\epsilon-hm)^2}$


We still need to analyse the stability of these equilibrium points but if the coexistence equilibrium is an attractor,
what can we say from the expression of these equilibria is:

- $P^*$ decreases with $a$ and $\epsilon$n and increases with $h$.

These equilibria are ecologically meaningfull when $P$ and $H$ are positive:
- Criterium for $P$ persistence: $\epsilon > hm$
- Criterium for $H$ persistence: $Kr_0a\epsilon > \frac{m}{1-hm}$



Further we can trace trajectories with different starting point to see to which equilibria they are attracted

```{r trace_isoclines_flowfield_with_trajectories}

# simulation time array
simtime = seq(0,1000,by = 0.1)     

#... Dynamic 1
yini1 = c(P=0.5,H=0.1)     #... initial conditions
dyn1 = data.frame(ode(y = yini1, times = simtime, func = rma_model, parms = pars, method = "rk4"))   #... run dynamics

#... Dynamics 2
yini2 = c(P=3,H=1)   #... initial conditions
dyn2 = data.frame(ode(y = yini2, times = simtime, func = rma_model, parms = pars, method = "rk4"))   #... run dynamics

# maximum values for plot scales
xmax = max(c(dyn1[,"P"],dyn2[,"P"]))*1.2
ymax = max(c(dyn1[,"H"],dyn2[,"H"]))*1.2


# plot isoclines
plot(NA,xlim = c(0,xmax),ylim = c(0,ymax),xlab="P",ylab="H",las=1, main="Phase plane\nisoclines, flow field and trajectories")
abline(h=0,col=colo[2],lwd=2)
abline(v=c(0,H_zerogrowth),col=colo,lwd=2)
lines(Ps[P_zerogrowth>=0],P_zerogrowth[P_zerogrowth>=0],col=colo[1],lwd=2)

# Add the flow field
library(phaseR)
flowfield = flowField(rma_model,xlim = c(0,xmax) ,ylim = c(0,ymax) , parameters = pars, points = 19, state.names = c("P","H"),add=T,las=1)

# Add the intersection points (equilibria)
points(alleqs[,"P"],alleqs[,"H"],lwd=2)

points(yini1[1],yini1[2],pch=16)           #... plot the initial point
lines(dyn1[,"P"],dyn1[,"H"])               #... plot the trajectory of the dynamics
points(dyn1[nrow(dyn1),"P"],dyn1[nrow(dyn1),"H"],pch=4,cex=2,lwd=3,col="red")  #...  endpoint


points(yini2[1],yini2[2],pch=16,col="purple")     #... plot the initial point
lines(dyn2[,"P"],dyn2[,"H"],col="purple",lty=2)   #... plot the trajectory of the dynamics
points(dyn2[nrow(dyn2),"P"],dyn2[nrow(dyn2),"H"],pch=4,cex=2,lwd=3,col="red")  #... endpoint



```

## Study the equilibria stability

To evaluate if these equilibrium are stable we need to evaluate the eigenvalues of the Jacobian matrix of the system at the specific equilibrium. The system is a stable node if the largest eigen value has a negative real part.


```{r local_stability1}

library(rootSolve)

# Initial values of the variables
yini = c(P=1,H=1) 
pars = c(r0 = 0.5, a = 0.9, eps = 0.5, m=0.5, h=0.4,K=7)

# Find the numerical steady-state (for a specific parameter set and initial densities)
Eq = stode(y = yini,func=rma_model,parms=pars,positive = TRUE)[[1]]


# Function with the differential equations to evaluate the jacobian matrix
jac <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dP = r0*P*(1-P/K) - a*P*H/(1+P*a*h)
    dH =  eps*a*P*H/(1+P*a*h) - m*H
    list(dP,dH)
  })
}

# Calculate the jacobian
J = jacobian.full(y=Eq,func=jac,parms=pars)


# Show the eigenvalues and the real part of the largest eigenvalue
eigen(J)$values
max(Re(eigen(J)$values))
```

For this parameter set, the dominant eignvalue has a negative real part, then the system is unstable. Actually it is an oscillating attractor.

## Bifuraction diagrams

A way to explore the model behavior is to plot the asymptotic (long-term) result along a gradient for a given parameter.
Because some asymptotic behavior can be oscillating, we record minima and maxima on a time window (which should be large enough to contin at least one complete oscillation) at the end of dynamics for each parameter set, while changind the value of the focal parameter:


```{r bifurcation_loop}

# Set the parameter values
r0 = 0.5     
K = 20
a=0.7
eps=0.3
m=0.4
h=0.4
pars = c(r0 = r0, a = a, eps = eps, m=m, h=h,K=K)

# Array of K values (Carrying capacity is our focal parameter)
Ks = seq(0.5,20,length=70)

simtime = seq(0,2000,by = 0.1)     # simulation time
yini = c("P"=1,"H"=1)              # initial densities

# initialize a dataframe to record the extrema along the K gradient
eqs = data.frame(P_min=NULL,P_max=NULL,H_min=NULL,H_max=NULL)

# loop to change K 
for(i in 1: length(Ks)){
  pars["K"] = Ks[i]
  dyn = data.frame(ode(y = yini, times = simtime, func = rma_model, parms = pars, method = "rk4"))
  dynend = dyn[(nrow(dyn)-1000):nrow(dyn),]
  eqs = rbind(eqs,data.frame(P_min=min(dynend[,"P"]),P_max=max(dynend[,"P"]),H_min=min(dynend[,"H"]),H_max=max(dynend[,"H"])))
}

# Takes a bit of time because we need to set a maximum time large enough to be sure we reach equilibria

```


Plot the resulting bifurcation diagram

```{r plot_bifurcation}
colos = rep(colo,each=2)
plot(NA,xlim = range(Ks),ylim=range(eqs),xlab="K",ylab="Extremum densities",las=1, main="Bifurcation diagram")
for(i in 1:4)points(Ks,eqs[,i],pch=16,col=colos[i])
legend("topleft",pch=16,col=colo,legend=c("P","H"),bty="n")

```
At low K, H is not present and P increases with K. At intermediate K, H survives and the system is stable. H regulates P and P density is constant and only depends on H's parameters. H densities benefits from increase in K. At higher K, the system is destabilizes and oscillates, due to a to large P population growth that is transferred with a delay to herbivores, that then explodes and overgraze the plants (paradox of enrichment sensu Rosenzweig 1971). 


The same bifurcation diagram can be done with the real part of the dominant eigenvalue of the Jacobian matrix to visualizing in another way the change in stability with K

```{r lambdamax_bifurcation}

lambdamax = NULL 


for(i in 1: length(Ks)){
  pars["K"] = Ks[i]
  Eq = stode(y = yini,func=rma_model,parms=pars,positive = TRUE)[[1]]
  J = jacobian.full(y=Eq,func=jac,parms=pars)
  lambdamax = c(lambdamax,max(Re(eigen(J)$values)))
}

plot(Ks,lambdamax,xlab="K",ylab="real part of dominant eigenvalue",pch=16,las=1,main="Bifurcation diagram, stability")
abline(h=0,lty=3)


```

## Multiple equilibria

Some models can display multiple attractors (asymptotic trajectories to which the system is attracted) either stable or unstable for the same parameter set. This means that, depending on the initial conditions, we will end up with different densities.

A quick example with the Lotka-Volterra competition model (NB: the Rosenzweig Mac-Arthur model doesn't display bistability:

$\frac{dN_1}{dt}=r_1N_1(1-\frac{N_1+\alpha_{21}N_2}{K_1}$
$\frac{dN_2}{dt}=r_2N_2(1-\frac{N_2+\alpha_{12}N_1}{K_2}$

with $\alpha_{21}$ the effect of species 2 on the growth rate of species 1


```{r bistable_dynamics}

# Lotka-Volterra 2-species competition model (function of derivatives)
lv_compet_ode <- function(t, y,parms) {
  with(as.list(c(y, parms)), {
    return(list(c(r1*y[1]*(1-(y[1]+a21*y[2])/K1),
                  r2*y[2]*(1-(y[2]+a12*y[1])/K2))))
  })
}

# Parameter set
pars=c(r1=0.1,r2=0.2,a12=1.5,a21=1.5,K1=15,K2=20)

# time array
simtime = seq(0,600,by = 0.1)

# first dynamics
init1 = c(5,5)                # initial densities
dyn1 = data.frame(ode(y = init1, times = simtime, func = lv_compet_ode, parms = pars, method = "rk4")) # run the dynamics

# second dynamics
init2 = c(10,1)               # initial densities
dyn2 = data.frame(ode(y = init2, times = simtime, func = lv_compet_ode, parms = pars, method = "rk4")) # run the dynamics

# Plot the dynamics for each initial conditions
layout(matrix(1:2,ncol=2))
coloN = c("darkorange","cyan3")

plot(dyn1[,1],dyn1[,2],xlab="times",ylim=range(dyn1[,2:3]),las=1,ylab="densities",main=paste("P0 =",init1[1],"H0 =",init1[2]),type="l",col=coloN[1],lwd=2)
lines(dyn1[,1],dyn1[,3],col=coloN[2],lwd=2)
legend("right",lwd=2,col=coloN,legend=c("N1","N2"),bty="n")

plot(dyn2[,1],dyn2[,2],xlab="times",ylim=range(dyn2[,2:3]),las=1,ylab="densities",main=paste("P0 =",init2[1],"H0 =",init2[2]),type="l",col=coloN[1],lwd=2)
lines(dyn2[,1],dyn2[,3],col=coloN[2],lwd=2)

```

### Findall equilibria with `nleqslv`

The principle is that we must run the dynamics for multiple initial conditions to find all the equilibria and then evaluate the stability of each to see if there are more than one stable equilibrium for the given parameter set.

```{r findallEq}
library(nleqslv)

# Lotka-Volterra model
lv_compet_model <- function(x, parms) {
  with(as.list(parms), {
    return(c(r1*x[1]*(1-(x[1]+a21*x[2])/K1),
             r2*x[2]*(1-(x[2]+a12*x[1])/K2)))
  })
}

# A matrix of initial densities from uniform distributions
yinimat = cbind(N1=runif(1000,min=0,max=20),N2=runif(100,min=0,max=20))

# Parameter set
pars=c(r1=0.1,r2=0.2,a12=1.5,a21=1.5,K1=15,K2=20)

# search all equilibria
eqs = searchZeros(yinimat,fn=lv_compet_model,parms=pars,control=list(xtol=10e-11,ftol=10e-11,allowSingular=TRUE))$x

# set to zero the near-zero values to exclude precision issue)
eqs = ifelse(eqs<0.0000001,0,eqs)   

## NB: searchZeros function can be useful if it is not possible to get the analytical equilibria (which is not the case here, so the code is just provided as an example), though precision seems to be some times problematic 


## Evaluate the stability of each equilibria we get
# formulation for jacobian
jac_lv <- function(t, x, parms) {
  with(as.list(c(x, parms)), {
    list(r1*x[1]*(1-(x[1]+a21*x[2])/K1),
         r2*x[2]*(1-(x[2]+a12*x[1])/K2))
  })
}

# Function to determine stability
stability_lv = function(Eq,pars){
  Jac = jacobian.full(y=Eq,func=jac_lv,parms=pars)     # calculate the Jacobian at equilibrium
  lambdamax = max(Re(eigen(Jac)$values))               # calculate the real part of the dominant eignevalue
  if(lambdamax < 0) return(1)                          # record stability (stable 1, unstable 0)
  else return(0)
}

# find the stability of each equilibrium
stab = numeric(nrow(eqs))
for(i in 1:nrow(eqs))stab[i] = stability_lv(eqs[i,],pars)
a = cbind(eqs,stab)

a
```



Analytical equilibria for the Lotka-Volterra two-species competition model:

$N_1^*=0, N_2^*=0$
$N_1^*=K_1, N_2^*=0$
$N_1^*=0, N_2^*=K_2$
$N_1^*=\frac{K_1-\alpha_{21}K_2}{1-\alpha_{21}\alpha_{12}}, N_2^*=\frac{K_2-\alpha_{12}K_1}{1-\alpha_{21}\alpha_{12}}$



```{r bifurcation_two_species_LV}

# Function to calculate equilibria from analytical formulas (2-column matrix with each row an equilibrium point)
lv_compet_eqs=function(pars){
  return(cbind(N1=c(0,pars["K1"],0,(pars["K1"]-pars["a21"]*pars["K2"])/(1-pars["a21"]*pars["a12"])),
               N2=c(0,0,pars["K2"],(pars["K2"]-pars["a12"]*pars["K1"])/(1-pars["a21"]*pars["a12"]))))
}


coloN = c("darkorange","cyan3")

# Number of values for the gradient
n=50

# Gradient of values for the interspecific competition coefficient a12
a12s = seq(0.5,1.5,length=n)


# parameter array
pars=c(r1=0.1,r2=0.2,a12=0.5,a21=1.5,K1=15,K2=20)

# object to record the equilibrium as well as the a12 values
res = c(a12 = NULL, N1=NULL,N2=NULL)

# For loop to change a12
for(i in 1:n){
  pars["a12"] = a12s[i]                                     # change the parameter value
  eqs = lv_compet_eqs(pars)                                 # calculate all equilibria
  stab = numeric(nrow(eqs))                                 # create an array to record stability
  for(j in 1:nrow(eqs))stab[j] = stability_lv(eqs[j,],pars) # assess equilibrium stability
  
  # record all the stable equilibria
  stable_eqs = eqs[which(stab==1),]
  if(length(stable_eqs) == 2){
    res = rbind(res, c(pars["a12"], c(stable_eqs["N1"]),stable_eqs["N2"]))
  }else{
    res = rbind(res, cbind(rep(pars["a12"],nrow(stable_eqs)),stable_eqs[,"N1"],stable_eqs[,"N2"]))
  }
}

# keep only the ones meeting feasibility criteria (densities > 0)
res_def =  res[which(res[,2]>=0 & res[,3]>=0 ),]

# plot all stable equilibrium states
plot(res_def[,"a12"],res_def[,"N1"],pch=16,col=coloN[1],ylim=c(0,max(res_def[,2:3])),xlab="a_12",ylab="equilibrium densities",las=1, main = "Multiple stable states\nLotka-Volterra 2-species competition")
points(res_def[,"a12"],res_def[,"N2"],pch=2,col=coloN[2])


```

For this parameter set, we have a zone of bistability at the end of the gradient where it's one or the other species excluding the second species.


```{r check_dynamics}
# Check dynamics for the first area
pars["a12"] = 1.4

yinit1 = c(20,1)
yinit2 = c(5,15)
dyn1 = data.frame(ode(y = yinit1, times = simtime, func = lv_compet_ode, parms = pars, method = "rk4"))
dyn2 = data.frame(ode(y = yinit2, times = simtime, func = lv_compet_ode, parms = pars, method = "rk4"))


# Plot the dynamics for each initial conditions
layout(matrix(1:2,ncol=2))
coloN = c("darkorange","cyan3")

plot(dyn1[,1],dyn1[,2],xlab="times",ylim=range(dyn1[,2:3]),las=1,ylab="densities",main=paste("P0 =",init1[1],"H0 =",init1[2]),type="l",col=coloN[1],lwd=2)
lines(dyn1[,1],dyn1[,3],col=coloN[2],lwd=2)
legend("right",lwd=2,col=coloN,legend=c("N1","N2"),bty="n")

plot(dyn2[,1],dyn2[,2],xlab="times",ylim=range(dyn2[,2:3]),las=1,ylab="densities",main=paste("P0 =",init2[1],"H0 =",init2[2]),type="l",col=coloN[1],lwd=2)
lines(dyn2[,1],dyn2[,3],col=coloN[2],lwd=2)

```



### 2-D parameter set exploration of stability and coexistence for the Rosenzweig-MacArthur model

In the following code we explore a 2D parameter space while varying the herbivore parameters *a* (grazing rate) and *h* (handling time)

```{r 2D_parameter_exploration}

col_coex = c("grey","chartreuse3","purple")

# Function to calculate equilibria from analytical formulas 
# (the output is a matrix with P and H the columns and the equilibria in rows)
rma_eqs=function(r0,K,a,h,eps,m){
  z = eps-h*m
  P=c(0,K,m/(a*eps-a*h*m))
  H=c(0,0,(eps*r0*(a*K*z-m))/(a^2*K*z^2))
  return(cbind(P,H))
}

# Set the parameter values
r0 = 0.5    
K = 20
a=0.7
eps=0.3
m=0.4
h=0.4

#... sequence of variation
n=30
a_s = seq(0.05,1.2,length=n)
h_s = seq(0.05,1.2,length=n)

#... object to record the parameters values, 
#... coexistence (0: none; 1: P only; 2: P, H), and stability (0: unstable; 1: stable)
res = c(h=NULL,a=NULL,coex=NULL,stab=NULL)

#... Nested loops to vary the parameters
for(a in a_s){
  for(h in h_s){
    eqs = rma_eqs(r0,K,a,h,eps,m)                            # calculate the equilibria
    pars = c(r0 = r0, a = a, eps = eps, m=m, h=h,K=K)       # parameter array
    
    # NB: we know that there is only one attractor for each parameter set in this model, either stable or oscillating 
    if(eqs[3,"P"]>0 & eqs[3,"H"]>0 ){                      # if the coexistence equilibrium is feasible (positive densities)
      JPH = jacobian.full(y=eqs[3,],func=jac,parms=pars) # calculate the Jacobian matrix at coexistence equilibrium
      stab = ifelse(max(Re(eigen(JPH)$value))<0,1,2)     # assess stability
      res = rbind(res,c(h=h,a=a,coex=3,stab=stab))       # record the results
    }else{
      JP = jacobian.full(y=eqs[2,],func=jac,parms=pars)    # if the coexistence equilibrium in not feasible, 
      # then only P persists. Calculate the jacobian for P alone 
      if(max(Re(eigen(JP)$value))<0){                      
        res = rbind(res,c(h=h,a=a,coex=2,stab=1))          # if P alone equilibrium is stable, then, record it
      }else{
        res = rbind(res,c(h=h,a=a,coex=1,stab=1))          # otherwise the stable equilibrium is necessary the last one
      }
    }
  }
} 


# Plot the results in the 2-D parameter space with colours coding coexistence
plot(res[,"h"],res[,"a"],pch=16,xlab="h",ylab="a",col=col_coex[res[,"coex"]],main="coexistence and stability",las=1)
# add white point where the dynamics is unstable
points(res[which(res[,"stab"]==2),"h"],res[which(res[,"stab"]==2),"a"],pch=16,cex=0.5,col="white")
legend("topright",pch=16,col=col_coex[c(2,3,3)],legend=c("P","PH stable","PH unstable"))
legend("topright",pch=16,col=c(col_coex[c(2,3)],"white"),pt.cex=c(1,1,0.5),legend=c("P","PH stable","PH unstable"),bty="n")

```

Look at one transect for $a=0.6$

```{r transect}

# Set the parameter values
r0 = 0.5     # r0 = 0.1
K = 20
a=0.6
eps=0.3
m=0.4
h=0.4

# parameter array


# gradient of h values
h_s = seq(0.05,1.2,length=n)

simtime = seq(0,2000,by = 0.1)     # simulation time
yini = c("P"=1,"H"=1)              # initial densities

# object to record results
eqs = data.frame(P_min=NULL,P_max=NULL,H_min=NULL,H_max=NULL)

for(h in h_s){
  pars = c(r0 = r0, a = a, eps = eps, m=m, h=h,K=K)        # parameter array
  dyn = data.frame(ode(y = yini, times = simtime, func = rma_model, parms = pars, method = "rk4")) # run the dynamic
  dynend = dyn[(nrow(dyn)-1000):nrow(dyn),]                # select the end of the dynamics (asymptotic state)
  
  # record extrema
  eqs = rbind(eqs,data.frame(P_min=min(dynend[,"P"]),P_max=max(dynend[,"P"]),H_min=min(dynend[,"H"]),H_max=max(dynend[,"H"])))
}

# Plot the results
colos = rep(colo,each=2)
plot(NA,xlim = range(h_s),ylim=range(eqs),xlab="h",ylab="Extremum densities",las=1, main = "Transect for a = 0.6")
for(i in 1:4)points(h_s,eqs[,i],pch=16,col=colos[i])
legend("topleft",pch=16,col=colo,legend=c("P","H"),bty="n")


```

## Model comparison

Compare the results of coexistence and stability in the $K$, $a$ parameter space for two models, one with a linear functional response and the other a saturating one (our previous Rsenzweig and MacArthur model).


```{r model_comparison}

final_state_types = c("none","P","P-H")
col_coex = c("grey","chartreuse3","purple")


# Set the parameter values
r0 = 0.5     # r0 = 0.1
K = 20
a=0.7
eps=0.3
m=0.4
h=0.4

#... sequence of variation
n=30
a_s = seq(0.05,1.2,length=n)
K_s = seq(0.05,20,length=n)

# Function to calculate equilibria from analytical formulas for the model with linear functional response
rma_linear_eqs=function(r0,K,a,eps,m){
  P=c(0,K,m/(a*eps))
  H=c(0,0,(eps*a*K-m)/(a^2*K*eps))
  return(cbind(P,H))
}


# Function to evaluate the jacobian matrix of the model with linear functional response
jac_linear <- function(t, y, parms) {
  with(as.list(c(y, parms)), {
    dP = r0*P*(1-P/K) - a*P*H
    dH =  eps*a*P*H - m*H
    list(dP,dH)
  })
}


# Objects to record the results for each model
res_linear = c(K=NULL,a=NULL,coex=NULL,stab=NULL)
res_nonlinear = c(K=NULL,a=NULL,coex=NULL,stab=NULL)


# Nested loops to vary K and a parameters
for(a in a_s){
  for(K in K_s){
    eqs_nonlinear = rma_eqs(r0,K,a,h,eps,m)
    eqs_linear = rma_linear_eqs(r0,K,a,eps,m)
    pars = c(r0 = r0, a = a, eps = eps, m=m, h=h,K=K)
    
    # Linear functional response model
    if(eqs_linear[3,"P"]>0 & eqs_linear[3,"H"]>0 ){
      JPH = jacobian.full(y=eqs_linear[3,],func=jac_linear,parms=pars)
      stab = ifelse(max(Re(eigen(JPH)$value))<0,1,2)
      res_linear = rbind(res_linear,c(K=K,a=a,coex=3,stab=stab))
    }else{
      JP = jacobian.full(y=eqs_linear[2,],func=jac_linear,parms=pars)
      if(max(Re(eigen(JP)$value))<0){
        res_linear = rbind(res_linear,c(K=K,a=a,coex=2,stab=1))
      }else{
        res_linear = rbind(res_linear,c(K=K,a=a,coex=1,stab=1))
      }
    }
    
    # Non linear functional response model
    if(eqs_nonlinear[3,"P"]>0 & eqs_nonlinear[3,"H"]>0 ){
      JPH = jacobian.full(y=eqs_nonlinear[3,],func=jac,parms=pars)
      stab = ifelse(max(Re(eigen(JPH)$value))<0,1,2)
      res_nonlinear = rbind(res_nonlinear,c(K=K,a=a,coex=3,stab=stab))
    }else{
      JP = jacobian.full(y=eqs_nonlinear[2,],func=jac,parms=pars)
      if(max(Re(eigen(JP)$value))<0){
        res_nonlinear = rbind(res_nonlinear,c(K=K,a=a,coex=2,stab=1))
      }else{
        res_nonlinear = rbind(res_nonlinear,c(K=K,a=a,coex=1,stab=1))
      }
    }
  }
} 

# Plot the results for each model
layout(matrix(1:2,ncol=2))

plot(res_linear[,"K"],res_linear[,"a"],pch=16,xlab="K",
     ylab="a",col=col_coex[res_linear[,"coex"]],main="Holling type I", las = 1)
points(res_linear[which(res_linear[,"stab"]==2),"K"],res_linear[which(res_linear[,"stab"]==2),"a"],
       pch=16,cex=0.5,col="white")

legend("topright",pch=16,col=col_coex[c(2,3,3)],legend=c("P","PH stable","PH unstable"),cex=0.7)
legend("topright",pch=16,col=c(col_coex[c(2,3)],"white"),pt.cex=c(1,1,0.5),legend=c("P","PH stable","PH unstable"),bty="n",cex=0.7)


plot(res_nonlinear[,"K"],res_nonlinear[,"a"],pch=16,xlab="K",
     ylab="a",col=col_coex[res_nonlinear[,"coex"]],main="Holling type II", las = 1)
points(res_nonlinear[which(res_nonlinear[,"stab"]==2),"K"],res_nonlinear[which(res_nonlinear[,"stab"]==2),"a"],
       pch=16,cex=0.5,col="white")




```




